---
title: "Statistics 652 - Midterm"
author: "Ashish Ashish"
date: today
format: 
  html:
    embed-resources: true
---

# Midterm

### Question 1

Clearly state the 5 steps for applying Machine Learning to a problem.

### **Answer:**

- Step 1:Define the problem and identify the response variable and potential predictors. Split the data into training and testing sets.

- Step 2:Perform exploratory data analysis and preprocess the training data such as handle missing values, create variables, scale predictors if needed.

- Step 3:Fit one or more candidate models using only the training data.

- Step 4:Evaluate and tune the models using the training data (e.g., cross-validation, accuracy, AUC, RMSE) and select the best-performing model.

- Step 5:Use the final trained model to make predictions on the testing data (which does not include the response variable) for final evaluation or submission.

### Question 2

For the [titanic](https://www.kaggle.com/c/titanic/data) data set try the following machine learning classification algorithms.

Use the training and test datasets from the **titanic** R package.

You should note that the *titanic_train* has the *Survived* variable and the *titanic_test* does not.  So to select your best model you need to use the *titanic_train* dataset to train and test your models.  So that means you will need to select a training dataset from *titanic_train* and select a testing dataset (this would be a validation dataset) from *titanic_train* to evaluate the models you try.

I have not demonstrated the use of cross-validation, once you are comfortable running all of the models see if you can figure out how to use cross-validation to pick the best model.

Once you have picked the best model you should do the following:

1. Re-run your chosen model on the full *titanic_train* dataset.
2. Then produce predictions for the *titanic_test* dataset.  This is what you would submit in a .csv to [Kaggle](https://www.kaggle.com/) in a competition.

Build **classification models** for the *Survived* variable. Pick an appropriate model scoring function, i.e. metric, and determine which model is the best. I would suggest making a confusion matrix and computing the accuracy or kappa. 

0. Null Model
1. kNN (the sample code given did not scale or normalize, if you use this model you need to do that.)
2. Boosted C5.0 
3. Random Forest
4. Logistic Regression using regularization
5. Naive Bayes

Extra Credit:

Make one plot containing all of the ROC curves for the algorithms trained.
```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
```


# Data
```{r}
library(pacman)
p_load(titanic, tidyverse, janitor, naniar, DataExplorer, tidymodels,kknn,C50,xgboost,ranger,glmnet,agua,klaR,yardstick,parsnip)
```
## Loading

```{r}
data(titanic_train)
data(titanic_test)
head(titanic_train)
head(titanic_test)
```

## Exploration

```{r}
library(naniar)
naniar::gg_miss_var(titanic_train)
naniar::gg_miss_var(titanic_test)
```

## Prep

```{r}
titanic_train <- titanic_train |>
  mutate(
    Survived = case_when(
      Survived %in% c(1, "1", "Yes", "yes") ~ "Yes",
      Survived %in% c(0, "0", "No",  "no")  ~ "No",
      TRUE ~ NA_character_
    ),
    Survived = factor(Survived, levels = c("No", "Yes")),
    Sex = as.factor(Sex),
    Embarked = as.factor(Embarked),
    Pclass = as.factor(Pclass),
    SibSp = as.double(SibSp),
    Parch = as.double(Parch)
  )

titanic_test <- titanic_test |>
    mutate(
        # Survived = factor(Survived, levels = c("0", "1"), labels = c("No", "Yes")),
        Sex = as.factor(Sex),
        Embarked = as.factor(Embarked),
        Pclass = as.factor(Pclass),
        SibSp = as.double(SibSp),
        Parch = as.double(Parch)
    )

head(titanic_train)
```


```{r}
# Training/validation split.
set.seed(108)
parts <- titanic_train |> initial_split(prop = 0.8, strata = Survived)
train <- parts |> training()
val <- parts |> testing()
```

```{r}
train <- train |> mutate(Survived = factor(Survived, levels = c("No","Yes")))
val   <- val   |> mutate(Survived = factor(Survived, levels = c("No","Yes")))
```


```{r}
rec <- recipe(Survived ~ ., data = train) |>
  update_role(PassengerId, Name, Ticket, Cabin, new_role = "id") |>
  step_rm(has_role("id")) |>
  step_impute_mean(all_numeric_predictors()) |>
  step_impute_mode(all_nominal_predictors()) |>
  step_novel(all_nominal_predictors()) |>
  step_zv(all_predictors())

preprocess <- rec

```


# Model 0 Null Model

### **Answer:**

The null model just predicts not survived, and has a validation accuracy of 61.5%.

```{r,warning=FALSE}
model_null <- null_model() |>
    set_engine("parsnip") |>
    set_mode("classification")

fit_null <- workflow() |>
    add_formula(Survived ~ .) |>
    add_model(model_null) |>
    last_fit(parts)

fit_null |> collect_metrics()

fit_null |> collect_predictions() |> conf_mat(truth = Survived, estimate = .pred_class)
```


# Model 1 kNN

### **Answer:**

or k nearest neighbors, after doing a grid search of k = 1 to 10 along with 10-fold cross validation, the optimal k value was found to be 7, yielding an accuracy of 78.7%.

The k nearest neighbors model found k = 8 to be the optimal v


```{r,warning=FALSE}
model_knn <- nearest_neighbor(neighbors = tune()) |>
    set_engine("kknn") |>
    set_mode("classification")

workflow_knn <- workflow() |>
    add_recipe(preprocess) |>
    add_model(model_knn)

fit_knn <- workflow_knn |>
    tune_grid(
        resamples = vfold_cv(train, v = 10, strata = Survived),
        grid = expand_grid(neighbors = 1:10),
        control = control_grid(save_pred = T)
    )

best_knn <- fit_knn |> select_best(metric = "accuracy")
best_knn

workflow_knn_final <- workflow_knn |> finalize_workflow(best_knn)

fit_knn_final <- workflow_knn_final |> last_fit(parts)

fit_knn_final |> collect_metrics()

fit_knn_final |>
    collect_predictions() |>
    conf_mat(truth = Survived, estimate = .pred_class)
```

# Model 2: Boosted C5.0

### **Answer:**

The Boosted C5.0 model obtained an accuracy of 79.8%.


```{r,warning=FALSE}
# Create a recipe just for C5.0
rec_c50 <- recipe(Survived ~ ., data = train) |>
  # Hard removal of ID columns prevents the C5.0 crash
  step_rm(PassengerId, Name, Ticket, Cabin) |>
  step_impute_mean(all_numeric_predictors()) |>
  step_impute_mode(all_nominal_predictors()) |>
  # Dummy variables ensure the C5.0 engine receives clean data
  step_dummy(all_nominal_predictors()) |>
  step_zv(all_predictors())

model_c50 <- boost_tree(trees = tune()) |>
    set_engine("C5.0") |>
    set_mode("classification")

workflow_c50 <- workflow() |>
    add_recipe(rec_c50) |> 
    add_model(model_c50)


fit_c50_tuned <- workflow_c50 |>
    tune_grid(
        resamples = vfold_cv(train, v = 10, strata = Survived),
        grid = expand_grid(trees = c(1, 10, 20, 30, 50)),
        control = control_grid(save_pred = TRUE)
    )

best_c50 <- fit_c50_tuned |> 
  select_best(metric = "accuracy")


workflow_c50_final <- workflow_c50 |> finalize_workflow(best_c50)


# This runs the 10-tree model on your training data and tests it
fit_c50_final <- workflow_c50_final |> last_fit(parts)


# This will show Accuracy and ROC AUC
fit_c50_final |> collect_metrics()

#  Confusion Matrix Table 
fit_c50_final |>
    collect_predictions() |>
    conf_mat(truth = Survived, estimate = .pred_class)
```



# Model 3: Random Forest


### **Answer:**

The random forest model has an accuracy of 81%.

```{r,warning=FALSE}
model_rf <- rand_forest(mtry = tune(), trees = tune(), min_n = tune()) |>
    set_engine("randomForest") |>
    set_mode("classification")

workflow_rf <- workflow() |>
    add_recipe(rec) |>
    add_model(model_rf)

fit_rf <- workflow_rf |>
    tune_grid(
        resamples = vfold_cv(train, v = 10, strata = Survived),
        grid = grid_latin_hypercube(
            mtry(range = c(1, 7)),
            trees(), min_n(),
            size = 10
        ),
        control = control_grid(save_pred = T)
    )

best_rf <- fit_rf |> select_best(metric = "accuracy")
best_rf

workflow_rf_final <- workflow_rf |> finalize_workflow(best_rf)

fit_rf_final <- workflow_rf_final |> last_fit(parts)

fit_rf_final |> collect_metrics()

fit_rf_final |>
    collect_predictions() |>
    conf_mat(truth = Survived, estimate = .pred_class)
```



# Model 4 Logistic Regression using regularization

### **Answer:**


The logistic regression model obtained an accuracy of 77.6%.

```{r,warning=FALSE}
model_lr <- logistic_reg(penalty = tune(), mixture = tune()) |>
    set_engine("glmnet") |>
    set_mode("classification")

# Extra preprocessing since logistic regression only works with numeric or dummy
# variables.
preprocess_lr <- rec |>
    step_dummy(Pclass, Sex, Embarked) 

workflow_lr <- workflow() |>
    add_recipe(preprocess_lr) |>
    add_model(model_lr)

fit_lr <- workflow_lr |>
    tune_grid(
        resamples = vfold_cv(train, v = 10, strata = Survived),
        grid = grid_latin_hypercube(penalty(), mixture(), size = 10),
        control = control_grid(save_pred = T)
    )

best_lr <- fit_lr |> select_best(metric = "accuracy")
best_lr

workflow_lr_final <- workflow_lr |> finalize_workflow(best_lr)

fit_lr_final <- workflow_lr_final |> last_fit(parts)

fit_lr_final |> collect_metrics()

fit_lr_final |>
    collect_predictions() |>
    conf_mat(truth = Survived, estimate = .pred_class)
```


# Model 5 Naive Bayes

### **Answer:**

The Naive Bayes model obtained an accuracy of 78.2%.

```{r,warning=FALSE}
# Required for Naive Bayes implementation.
library(agua)
library(discrim)

model_nb <- naive_Bayes(smoothness = tune(), Laplace = tune()) |>
    set_engine("klaR") |>
    set_mode("classification")

workflow_nb <- workflow() |>
    add_recipe(preprocess) |>
    add_model(model_nb)

fit_nb <- workflow_nb |>
    tune_grid(
        resamples = vfold_cv(train, v = 10, strata = Survived),
        grid = grid_latin_hypercube(smoothness(), Laplace(), size = 10),
        control = control_grid(save_pred = T)
    )

best_nb <- fit_nb |> select_best(metric = "accuracy")
best_nb

workflow_nb_final <- workflow_nb |> finalize_workflow(best_nb)

fit_nb_final <- workflow_nb_final |> last_fit(parts)

fit_nb_final |> collect_metrics()

fit_nb_final |>
    collect_predictions() |>
    conf_mat(truth = Survived, estimate = .pred_class)
```


# Final

The model with the highest accuracy (81%) was the random forest model, so that will be used to fit on the entire training set.

**There is an error in the next code chunk.  Needs to be updated.**
```{r}
levels(titanic_train$Survived)
```

```{r}
# Finalize RF model with best parameters
model_final <- model_rf |> finalize_model(best_rf)

workflow_final <- workflow() |>
  add_recipe(preprocess) |>
  add_model(model_final)

# Fit on full training data
fit_final <- workflow_final |> fit(data = titanic_train)

# Predict on test data (NO bake needed)
pred_final <- fit_final |> predict(new_data = titanic_test)

# Kaggle submission format (Yes/No -> 1/0)
submission_rf <- titanic_test |>
  dplyr::select(PassengerId) |>
  mutate(Survived = ifelse(pred_final$.pred_class == "Yes", 1, 0))

head(submission_rf,20)
```

```{r}
val_yesno <- val |>
  mutate(
    Survived = as.character(Survived),
    Survived = ifelse(Survived == "Yes", "Yes", "No"),
    Survived = factor(Survived, levels = c("No","Yes"))
  )

val_01 <- val |>
  mutate(
    Survived = as.character(Survived),
    Survived = ifelse(Survived == "Yes", "1", "0"),
    Survived = factor(Survived, levels = c("0","1"))
  )

# MUST show both 0 and 1
table(val_yesno$Survived)
table(val_01$Survived)
```
```{r}
# val already has No/Yes; create matching 0/1 truth for the other models
val_01 <- val |>
  mutate(
    Survived = ifelse(Survived == "Yes", "1", "0"),
    Survived = factor(Survived, levels = c("0","1"))
  )

roc_data <- bind_rows(
  # Null
  extract_workflow(fit_null) |>
    predict(val_01, type = "prob") |>
    bind_cols(val_01) |>
    roc_curve(truth = Survived, .pred_Yes, event_level = "second") |>
    mutate(model = "Null"),

  # kNN
  extract_workflow(fit_knn_final) |>
    predict(val_01, type = "prob") |>
    bind_cols(val_01) |>
    roc_curve(truth = Survived, .pred_Yes, event_level = "second") |>
    mutate(model = "kNN"),

  # Random Forest
  extract_workflow(fit_rf_final) |>
    predict(val_01, type = "prob") |>
    bind_cols(val_01) |>
    roc_curve(truth = Survived, .pred_Yes, event_level = "second") |>
    mutate(model = "Random Forest"),

  # Logistic Regression
  extract_workflow(fit_lr_final) |>
    predict(val_01, type = "prob") |>
    bind_cols(val_01) |>
    roc_curve(truth = Survived, .pred_Yes, event_level = "second") |>
    mutate(model = "Logistic Regression"),

  # Naive Bayes
  extract_workflow(fit_nb_final) |>
    predict(val_01, type = "prob") |>
    bind_cols(val_01) |>
    roc_curve(truth = Survived, .pred_Yes, event_level = "second") |>
    mutate(model = "Naive Bayes"),

  # Boosted C5.0
  extract_workflow(fit_c50_final) |>
    predict(val, type = "prob") |>
    bind_cols(val) |>
    roc_curve(truth = Survived, .pred_Yes, event_level = "second") |>
    mutate(model = "Boosted C5.0")
)
```
```{r}
ggplot(roc_data, aes(x = 1 - specificity, y = sensitivity, color = model)) +
  geom_path(linewidth = 1.1) +
  geom_abline(lty = 3) +
  coord_equal() +
  theme_minimal()
```

